\LARGE{\bf Parallelizing a Network Simulator}

### 2009-03-17 doc/disc.tex
Second revision of report.
git-svn-id: https://nocs.stanford.edu/cgi-bin/svn.cgi/booksim2.0/tags/pbooksim@1159 ce7aba0d-9f81-40a6-a495-cecae84b0515

\subsection{Static Network Partition}
The current implementation divides the network into equal partitions among the threads and seeks to minimize computation. This organization works well when the traffic pattern is balanced across the partitions. However, hotspot traffic patterns which concentrate the packets into a particular section of the network will cause imbalances in the amount of work handled by each thread. Even with the shared channel semaphore and the removal of the cycle barrier, the lightly loaded threads will  always be stalled waiting for the hotspot threads. Fortunately, common network traffic patterns are balanced and network routing algorithm are usually designed to reduce hotspots and bottlenecks in the network. Nevertheless, it is not a guarantee that an equally partitioned network will give equal work to all threads. Therefore, future improvements for the simulator should include dynamic partitioning of the network among the threads. This can be implemented by changing the current loop based iteration of network components into a work queue based organization. A work queue can be created for each network partition and each queue entry record the network component to be operated on and the network cycle associated with the operation. After operating on a component, the component is reinserted into the queue and the associated cycle is incremented. Under this organization, threads that have stalled ahead can pull work from other threads' work queues and help to finish the work in other network partitions.

\subsection{Programming Model}
Given the underwhelming speedup that we obtained through parallelization and the possibility of excessive false sharing in the program, this has lead us to think that maybe PThreads is not the most appropriate programming model for this simulator. As we have stated through out the report, the only points of sharing between threads are the channels that bridge routers running under different threads. No other source of true sharing should exist in the simulator. But with the shared memory model of PThreads, false sharing due to the organization of data structure can easily occur. Since we know the exact points of data sharing and since the smart partitioning of the networks should keep the amount of inter-thread communication small, perhaps message passing programming models such as MPI will offer more scalability for the parallel simulator. Shared channel objects can be written on top of MPI's send and receive interfaces, and many packets can be bundled together to reduce messaging overhead.

>>>>

\subsection{Static Network Partition}
The current implementation divides the network into equal partitions among the threads and seeks to minimize computation. This organization works well when the traffic pattern is balanced across the partitions. However, hotspot traffic patterns which concentrate the packets into a particular section of the network will cause imbalances in the amount of work handled by each thread. Even with the shared channel semaphore and the removal of the cycle barrier, the lightly loaded threads will  always be stalled waiting for the hotspot threads. Fortunately, common network traffic patterns are balanced and network routing algorithm are usually designed to reduce hotspots and bottlenecks in the network. Nevertheless, it is not a guarantee that an equally partitioned network will give equal work to all threads. Therefore, future improvements for the simulator should include dynamic partitioning of the network among the threads. This can be implemented by changing the current loop based iteration of network components into a work queue based organization. A work queue can be created for each network partition and each queue entry record the network component to be operated on and the network cycle associated with the operation. After operating on a component, the component is reinserted into the queue and the associated cycle is incremented. Under this organization, threads that have stalled ahead can pull work from other threads' work queues and help to finish the work in other network time slices.

\subsection{Programming Model}
Given the underwhelming speedup that we obtained through parallelization and the possibility of excessive false sharing in the program, this has lead us to think that PThreads is not the most appropriate programming model for this simulator. As we have stated through out the report, the only points of sharing between threads are the channels that bridge routers running under different threads. No other source of true sharing should exist in the simulator. But with the shared memory model of PThreads, false sharing due to the organization of data structure can easily occur. Since we know the exact points of data sharing and since the smart partitioning of the networks should keep the amount of inter-thread communication small, perhaps message passing programming models such as MPI could offer more scalability for the parallel simulator. Shared channel objects can be written on top of MPI's send and receive interfaces, and many packets can be bundled together to reduce messaging overhead.

### 2009-03-17 doc/paper.tex
Second revision of report.
git-svn-id: https://nocs.stanford.edu/cgi-bin/svn.cgi/booksim2.0/tags/pbooksim@1159 ce7aba0d-9f81-40a6-a495-cecae84b0515

multithreaded programs.  While this has greatly increased the computation power of the processors, existing programs that have been

@@

of data. The traditional bus architecture is no longer the appropriate medium of communication for large number of cores. Even more recently, single crossbar communication hubs will also not be sufficient in the future since the cost of a single crossbar grows quadratically with the number of cores. Interconnection network can provide the scalable and efficient communication

@@

\section{Parallel Implementation \label{para}}
In order to parallelize the network simulator, we first set about
identifying sources of parallelism that we could exploit.  As mentioned previous, the routers in a network operate relatively autonomously.   Each router in the network can be simulated
independently as long as the communication that occurs between the routers
is synchronized in the channels.  In this section we describe how we
partitioned the work of the network simulator and the synchronization
mechanisms that we employed in order to maintain the correctness of the simulator.

@@

\subsubsection{Locking Channels}
We initially assigned routers to threads by their ID numbers which are completely independent of
topology.  This left us no easy way of determining which channels would
ultimately be shared and which would be local to a thread, therefore we
were forced to place locks on all channels to ensure correctness.  Since
the channels were initially represented as STL queues, there was no way to
allow concurrent accesses to the queues (i.e. we couldn't allow one router
to be reading and another to be writing simultaneously).  In order to
prevent this, we protected each queue with a PThread lock.  For a thread
to access the queue, it first had to acquire the lock and then release the
lock when finished performing its operation on the queue.\\
~\\
While this approach guaranteed correctness, it also caused significant
performance overhead.  Even queues that were not shared between different
threads were having to be locked and unlocked in order to perform
operations on the queues.  In order to get around this additional
overhead, we instead began to partition the router work based upon the
topology of the network.  This allowed us to know which channels
were shared and which were local to a single thread.  By doing this, we
were able to remove some of the overhead of locking when performing
operations on a channel's queue. However, this method adds an additional layer of complexity for the user. For each topology in the simulator, we need to write a network partitioning algorithm for that topology. Ideally we would like a generalized algorithm that would discover the optimal partitioning to minimize locking given any network, but this is not within the scope of this project and is a very difficult (NP-Complete) problem by itself.

@@

\section{Results \label{results}}
Having completed a parallel implementation of our network simulator, we first tested its functionality. We generated a series of latency-throughput plot for a dragonfly network running with different traffic patterns. The resulting plot was nearly identical to those of the sequential simulator. On average there was about less than 0.1\% difference between the two versions. The slight different was caused by the randomness of packet injection. \\
~\\
 Our next step was to analyze the performance of the simulator to determine how
much speedup it achieved.  We began first by analyzing the performance of
the different synchronization mechanisms to ensure that they held with our
hypotheses about which method permitted the most concurrency.  We then analyzed
our results on a series of architectures to determine the scalability of
the simulator as well as any potential bottlenecks.

\subsection{Analyzing Synchronization Mechanisms}
The first step in analyzing our parallelized network simulator was to
determine the performance of the different synchronization mechanisms that
we had employed at different points in development.  Figure \ref{synch}
shows the performance of the various synchronization mechanisms on a
single node of the Hotbox cluster that consists of two Intel Xeon Quad
cores.  We only ran experiments up to 8 threads as this is the maximum
number of hardware thread contexts available without encountering the
overhead of using ethernet to communicate between nodes.\\
\begin{figure}[h]
\centering
\includegraphics[width=3in,height=3in]{synch.eps}
\caption{\label{synch} This figure shows the performance of the different
synchronization mechanisms on up to eight threads on the Hotbox cluster.}
\end{figure}
~\\
From these results we can see that our hypotheses were confirmed regarding
the performance of the various synchronization mechanisms.  Barriering
every cycle added significant overhead to performance and ultimately
prevented even 8 threads from extracting much parallelism at all.
Surprisingly the same result was discovered when the barrier was removed and
all the queues were replaced with lock-free queues.  However, once we were
able to only implement the shared channels with lock-free queues , we could see a noticeable performance gain and even some
additional scalability.  Finally, once we fused several of the loops being
executed within the program, we were able to improve some of the temporal
locality of data use and thereby achieve a little additional
performance.\\
~\\
Despite our efforts however, we were only able to extract a two fold
speedup using eight threads, far below a linear speedup.  This was very
perplexing to us as we had consciously attempted to break up the work into
parallel computations that required very little inter-thread
communication.  In order to determine exactly what was causing us to lose
performance we then proceeded to run several experiments to find the
bottlenecks in the program.

@@

Our initial hypothesis was that application was memory bound based on the

@@

cache, then there will be a significant amount of data being moved both to

@@

The first method of our program analysis was by using Intel's Vtune performance analyzer \cite{vtune}. Vtune provides a thread profiler for characterizing multi-threaded programs as well as access to all performance counters in Intel processors. The comparison between the profile of the parallel code and that of the sequential code did not reveal much insights. Majority of the execution time is still within the routing step during each simulation cycle. The lock free queues account for about 10\% of the runtime, proving that synchronization is a performance problem. One performance counter that did stand out is the data bus utilization. The utilization rate for the sequential code is 35\% while the parallel code with 2 threads was at 50\%. After tallying the number of cache lines transferred on the data bus (the BUS\_TRANS\_BURST.ALL\_AGENT counter) and use the formula provided by \cite{account}, we calculated that the parallel program uses approximated 3GB/s memory bandwidth from the front side bus. The maximum possible bandwidth on the system is 6.4GB/s, this was a further indication to us that simulator is perhaps memory bound. 

@@

\ref{niagara}.

@@

began to interfere with each other in the shared L2 cache is when the
performance begins to degrade and couldn't offer us much insight into  any potential false sharing that was occurring when run on other

>>>>

multithreaded programs.  While this has greatly increased the computational power of the processors, existing programs that have been

@@

of data. The traditional bus architecture is no longer the appropriate medium of communication for a large number of cores. Even more recently, single crossbar communication hubs will also not be sufficient in the future since the cost of a single crossbar grows quadratically with the number of cores. Interconnection networks can provide the scalable and efficient communication

@@

\section{Parallel Implementation \label{para}}
In order to parallelize the network simulator, we first set about
identifying sources of parallelism that we could exploit.  As mentioned previously, the routers in a network operate relatively autonomously.   Each router in the network can be simulated
independently as long as the communication that occurs between the routers
is synchronized in the channels.  In this section we describe how we
partitioned the work of the network simulator and the synchronization
mechanisms that we employed in order to maintain the correctness of the simulator.

@@

\subsubsection{Locking Channels}
We initially assigned routers to threads by their ID numbers which are completely independent of
topology.  This left us no easy way of determining which channels would
ultimately be shared and which would be local to a thread, therefore we
were forced to place locks on all channels to ensure correctness.  Since
the channels were initially represented as STL queues, there was no way to
allow concurrent accesses to the queues (i.e. we couldn't allow one router
to be reading and another to be writing simultaneously).  In order to
prevent this, we protected each queue with a PThread lock.  For a thread
to access the queue, it first had to acquire the lock and then release the
lock when it finished performing its operation on the queue.\\
~\\
While this approach guaranteed correctness, it also caused significant
performance overhead.  Even queues that were not shared between different
threads were having to be locked and unlocked in order to perform
operations on the queues.  In order to get around this additional
overhead, we instead began to partition the router work based upon the
topology of the network.  This allowed us to know which channels
were shared and which were local to a single thread.  By doing this, we
were able to remove some of the overhead of locking when performing
operations on a channel's queue. However, this method adds an additional layer of complexity for the user. For each topology in the simulator, we needed to write a network partitioning algorithm for that topology. Ideally we would like a generalized algorithm that would discover the optimal partitioning to minimize locking given any network, but this is not within the scope of this project and is a very difficult (NP-Complete) problem by itself.

@@

\section{Results \label{results}}
Having completed a parallel implementation of our network simulator, we first tested its functionality. We generated a series of latency-throughput plots for a dragonfly network running with different traffic patterns. The resulting plot was nearly identical to those of the sequential simulator. On average there was less than 0.1\% difference between the two versions. The slight different was caused by the randomness of packet injection. \\
~\\
 Our next step was to analyze the performance of the simulator to determine how
much speedup it achieved.  We began first by analyzing the performance of
the different synchronization mechanisms to ensure that they held with our
hypotheses about which method permitted the most concurrency.  We then analyzed
our results on a series of architectures to determine the scalability of
the simulator as well as any potential bottlenecks.

\subsection{Analyzing Synchronization Mechanisms}
The first step in analyzing our parallelized network simulator was to
determine the performance of the different synchronization mechanisms that
we had employed at different points in development.  Figure \ref{synch}
shows the performance of the various synchronization mechanisms on a
single node of the Hotbox cluster that consists of two Intel Xeon Quad
cores.  We only ran experiments up to 8 threads as this is the maximum
number of hardware thread contexts available without encountering the
overhead of using ethernet to communicate between nodes.\\
\begin{figure}[h]
\centering
\includegraphics[width=3in,height=3in]{synch.eps}
\caption{\label{synch} This figure shows the performance of the different
synchronization mechanisms on up to eight threads on the Hotbox cluster.}
\end{figure}
~\\
From these results we can see that our hypotheses were confirmed regarding
the performance of the various synchronization mechanisms.  Barriering
every cycle added significant overhead to performance and ultimately
prevented even 8 threads from extracting much parallelism at all.
Surprisingly the same result was discovered when the barrier was removed and
all the queues were replaced with lock-free queues.  However, once we were
able to only implement the shared channels with lock-free queues, we could see a noticeable performance gain and even some
additional scalability.  Finally, once we fused several of the loops being
executed within the program, we were able to improve some of the temporal
locality of data use and thereby achieve a little additional
performance.\\
~\\
Despite our efforts however, we were only able to extract a two fold
speedup using eight threads, far below a linear speedup.  This was very
perplexing to us as we had consciously attempted to break up the work into
parallel computations that required very little inter-thread
communication.  In order to determine exactly what was causing us to lose
performance we then proceeded to run several experiments to find the
bottlenecks in the program.

@@

Our initial hypothesis was that the application was memory bound based on the

@@

cache, then there would be a significant amount of data being moved both to

@@

The first method of our program analysis was to use Intel's Vtune performance analyzer \cite{vtune}. Vtune provides a thread profiler for characterizing multi-threaded programs as well as access to all performance counters in Intel processors. The comparison between the profile of the parallel code and that of the sequential code did not reveal much insights. The majority of the execution time was still within the routing step during each simulation cycle. The lock free queues account for about 10\% of the runtime, proving that synchronization is still a performance problem. One performance counter that did stand out is the data bus utilization. The utilization rate for the sequential code was 35\% while the parallel code with 2 threads was at 50\%. After tallying the number of cache lines transferred on the data bus (the BUS\_TRANS\_BURST.ALL\_AGENT counter) and use the formula provided by \cite{account}, we calculated that the parallel program used approximated 3GB/s memory bandwidth from the front side bus. The maximum possible bandwidth on the system is 6.4GB/s, this was a further indication to us that simulator was perhaps memory bound. \\

@@

\ref{niagara}.\\

@@

began to interfere with each other in the shared L2 cache was when the
performance began to degrade and couldn't offer us much insight into any potential false sharing that was occurring when run on other

### 2009-03-17 doc/sim.tex
Second revision of report.
git-svn-id: https://nocs.stanford.edu/cgi-bin/svn.cgi/booksim2.0/tags/pbooksim@1159 ce7aba0d-9f81-40a6-a495-cecae84b0515

The interconnection network simulator, Booksim, was created by the Concurrent VLSI Architecture group (CVA) at Stanford to study network on chip and large scale computer systems. The simulator allows for a wide range of user customizations when constructing a network. The simulator supports a variety of network topologies, router micro-architectures, routing algorithms, traffic pattern, and simulation modes. The priority of the simulator is  to guarantee the correctness and accuracy of simulation, and as a result the simulator is implemented in a cycle-accurate fashion. Booksim has been used to generate results for many research publications \cite{cdragon,KimISCA07,PPIN}. 

@@

\caption{Simulator organization: Network is composed of routers and channels and simulations are run through the traffic manager. }

@@

Figure~\ref{fig:runtime} diagrams the runtime steps of the simulator. At the highest level is the network cycle loop. Each iteration of this loop represents a cycle in the network. Within each network cycle, several steps occurs to advance the state of the network. At each step, the simulator loops through all the routers in the network to perform a specific function. As mentioned previously the operation of routers are mostly independent of each other. Therefore, there are no dependencies between each iterations of these loops. Channels are the only source of sharing between the routers 

>>>>

The interconnection network simulator, Booksim, was created by the Concurrent VLSI Architecture group (CVA) at Stanford to study network on chip and large scale computer systems. The simulator allows for a wide range of user customizations when constructing a network. The simulator supports a variety of network topologies, router micro-architectures, routing algorithms, traffic patterns, and simulation modes. The priority of the simulator is  to guarantee the correctness and accuracy of simulation, and as a result the simulator is implemented in a cycle-accurate fashion. Booksim has been used to generate results for many research publications \cite{cdragon,KimISCA07,PPIN}. 

@@

\caption{Simulator organization: the network is composed of routers and channels and simulations are run through the traffic manager. }

@@

Figure~\ref{fig:runtime} diagrams the runtime steps of the simulator. At the highest level is the network cycle loop. Each iteration of this loop represents a cycle in the network. Within each network cycle, several steps occur to advance the state of the network. At each step, the simulator loops through all the routers in the network to perform a specific function. As mentioned previously the operation of a router is mostly independent of other routers. Therefore, there are no dependencies between each iteration of these loops. Channels are the only source of sharing between the routers 
